From Reversible Programs to \\ Univalent Universes and Back

### 2017-05-12 upi.tex
univalent fibration non-example

We will consider a general statement of \emph{univalence} for type families. We
start with a some preliminaries. Fig.~\ref{fig:fib} depicts a general type
family $P : A \rightarrow \mathcal{U}$ mapping each point $x : A$ to a space
$P(x)$. As everything is continuous, we expect this type family to respect
identities. In other words if $x:A$ and $y:A$ are identified, it must be the
case that the spaces $P(x)$ and $P(y)$ are ``equivalent.''

\begin{figure}
\begin{tabular}{c@{\qquad\qquad}c}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $A$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$x$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$y$};
  \draw (-1,-2) ellipse (0.5cm and 2cm);
  \node[left] at (-1.5,-2) {Fiber $P(x)$};
  \draw (1,-2) ellipse (0.5cm and 2cm);
  \node[right] at (1.5,-2) {Fiber $P(y)$};
\end{tikzpicture}
&
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $A$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$x$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$y$};
  \draw (-1.3,-2) ellipse (0.5cm and 2cm);
  \node[left] at (-1.8,-2) {Fiber $P(x)$};
  \draw (1.3,-2) ellipse (0.5cm and 2cm);
  \node[right] at (1.8,-2) {Fiber $P(y)$};
  \draw[below,cyan,thick] (-1,-5) -- (1,-5);
  \node[below,cyan,thick] at (0,-5) {$p$};
  \draw[->,red,ultra thick] (-0.8,-1.7) to [out=45, in=135] (0.8,-1.7);
  \draw[->,red,ultra thick] (0.8,-2.3) to [out=-135, in=-45] (-0.8,-2.3);
  \node[red,ultra thick] at (0,-2) {$\simeq$};
\end{tikzpicture}
\end{tabular}
\caption{\label{fig:fib}(left)Type family $P : A \rightarrow \mathcal{U}$ as a
  fibration with total space $\Sigma_{(x:A)} P(x)$; (right) a path $x \equiv y$
  induces an equivalence between the spaces $P(x)$ and $P(y)$}
\end{figure}

The converse statement is whether an equivalence between $P(x)$ and $P(y)$
necessarily induces an identification $x \equiv y$. It is easy to construct
several examples in which this converse statement fails and it instructive to
examine such examples.

>>>>

We will consider a general statement of \emph{univalence} for type families. We
start with a some preliminaries. Fig.~\ref{fig:fib} depicts a type family
$P : A \rightarrow \mathcal{U}$ mapping each point $x : A$ to a space $P(x)$. As
everything is continuous, we expect this type family to respect identities. In
other words if $x:A$ and $y:A$ are identified using a path $x \equiv y$, it must
be the case that the spaces $P(x)$ and $P(y)$ become equivalent
$P(x) \simeq P(y)$ as a consequence. Generally speaking, there is however no
reason to expect that the spaces $x \equiv y$ and $P(x) \simeq P(y)$ to be
themselves equivalent, and indeed we will provide a few examples in which they
are not. When they are indeed equivalent, the fibration is a \emph{univalent
  fibration}.

\begin{figure}
\begin{tabular}{c@{\qquad\qquad}c}
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $A$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$x$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$y$};
  \draw (-1,-2) ellipse (0.5cm and 2cm);
  \node[left] at (-1.5,-2) {Fiber $P(x)$};
  \draw (1,-2) ellipse (0.5cm and 2cm);
  \node[right] at (1.5,-2) {Fiber $P(y)$};
\end{tikzpicture}
&
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $A$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$x$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$y$};
  \draw (-1.3,-2) ellipse (0.5cm and 2cm);
  \node[left] at (-1.8,-2) {Fiber $P(x)$};
  \draw (1.3,-2) ellipse (0.5cm and 2cm);
  \node[right] at (1.8,-2) {Fiber $P(y)$};
  \draw[below,cyan,thick] (-1,-5) -- (1,-5);
  \node[below,cyan,thick] at (0,-5) {$\equiv$};
  \draw[->,red,ultra thick] (-0.8,-1.7) to [out=45, in=135] (0.8,-1.7);
  \draw[->,red,ultra thick] (0.8,-2.3) to [out=-135, in=-45] (-0.8,-2.3);
  \node[red,ultra thick] at (0,-2) {$\simeq$};
\end{tikzpicture}
\end{tabular}
\caption{\label{fig:fib}(left)Type family $P : A \rightarrow \mathcal{U}$ as a
  fibration with total space $\Sigma_{(x:A)} P(x)$; (right) a path $x \equiv y$
  induces an equivalence between the spaces $P(x)$ and $P(y)$}
\end{figure}

As a simple example of a non-univalent fibration consider $P : \mathbb{2}
\rightarrow \mathbb{U}$ defined as $P = \lambda \_. \mathbb{1}$ illustrated below:

\bigskip
\begin{tikzpicture}[scale=0.7,every node/.style={scale=0.7}]]
  \draw (0,-5) ellipse (2cm and 0.8cm);
  \node[below] at (0,-6) {Base Space $\mathbb{2}$};
  \draw[fill] (-1,-5) circle [radius=0.025];
  \node[below] at (-1,-5) {$\fc$};
  \draw[fill] (1,-5) circle [radius=0.025];
  \node[below] at (1,-5) {$\tc$};
  \draw (0,-2.7) ellipse (0.5cm and 1cm);
  \node[right] at (0.5,-2.7) {Fiber $P(\fc)$ = Fiber $P(\tc)$ = $\mathbb{1}$};
  \draw[fill] (0,-2.7) circle [radius=0.025];
  \node[below] at (0,-2.7) {$\star$};
\end{tikzpicture}

\medskip\noindent There is one equivalence
$P(\fc) = \mathbb{1} \simeq \mathbb{1} = P(\tc)$ but this certainly does not
induce a path $\fc \equiv \tc$.
